---
title: "Feature Engineering: Affordability Analysis"
author: "Harjot Singh "
date: "2026-05-03"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
  pdf_document: default
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

options(scipen = 999)
```

# Purpose

This file completes the feature engineering step for the housing prediction project.

For this step, we are ignoring regional indicators for now because the current datasets are statewide California data. Instead, this file focuses on creating affordability-related features using:

- House Price Index
- Median household income
- Mortgage interest rates

Important note: HPI is an index, not an actual dollar home price. Because of that, the HPI-to-income ratio should be interpreted as a relative affordability indicator, not the literal number of years of income needed to buy a house.

# Load Data

```{r load-data}
required_files <- c(
  "CASTHPI_yearly_cleaned.csv",
  "cleaned_income_data.csv",
  "cleaned_mortgage_rates_annual_avg.csv"
)

missing_files <- required_files[!file.exists(required_files)]

if (length(missing_files) > 0) {
  stop(
    paste(
      "Missing file(s):",
      paste(missing_files, collapse = ", "),
      "\nCurrent working directory is:",
      getwd(),
      "\nMove this Rmd file into the same folder as the CSV files or set the working directory correctly."
    )
  )
}

hpi_data <- read_csv("CASTHPI_yearly_cleaned.csv")

income_data <- read_csv("cleaned_income_data.csv") %>%
  rename(Year = Date)

mortgage_data <- read_csv(
  "cleaned_mortgage_rates_annual_avg.csv",
  skip = 7,
  col_names = c(
    "Year",
    "Average_Mortgage_Rate",
    "Lowest_Mortgage_Rate",
    "Highest_Mortgage_Rate",
    "Number_Of_Weeks"
  )
)

glimpse(hpi_data)
glimpse(income_data)
glimpse(mortgage_data)
```

# Clean and Merge Data

```{r clean-and-merge}
hpi_data <- hpi_data %>%
  mutate(
    Year = as.numeric(Year),
    Yearly_Average_HPI = as.numeric(Yearly_Average_HPI),
    Lowest_Quarterly_HPI = as.numeric(Lowest_Quarterly_HPI),
    Highest_Quarterly_HPI = as.numeric(Highest_Quarterly_HPI)
  )

income_data <- income_data %>%
  mutate(
    Year = as.numeric(Year),
    Median_Household_Income = as.numeric(Median_Household_Income)
  )

mortgage_data <- mortgage_data %>%
  mutate(
    Year = as.numeric(Year),
    Average_Mortgage_Rate = as.numeric(Average_Mortgage_Rate),
    Lowest_Mortgage_Rate = as.numeric(Lowest_Mortgage_Rate),
    Highest_Mortgage_Rate = as.numeric(Highest_Mortgage_Rate),
    Number_Of_Weeks = as.numeric(Number_Of_Weeks)
  )

combined_data <- hpi_data %>%
  left_join(income_data, by = "Year") %>%
  left_join(mortgage_data, by = "Year") %>%
  arrange(Year) %>%
  drop_na()

glimpse(combined_data)

knitr::kable(
  head(combined_data),
  caption = "Preview of merged HPI, income, and mortgage data"
)
```

# Create New Features

```{r feature-engineering}
feature_data <- combined_data %>%
  mutate(
    HPI_to_Income_Ratio = Yearly_Average_HPI / Median_Household_Income,

    HPI_per_100k_Income = Yearly_Average_HPI / (Median_Household_Income / 100000),

    Affordability_Index = HPI_to_Income_Ratio / first(HPI_to_Income_Ratio),

    Income_Growth_From_Start = Median_Household_Income / first(Median_Household_Income),

    HPI_Growth_From_Start = Yearly_Average_HPI / first(Yearly_Average_HPI),

    Affordability_Gap = HPI_Growth_From_Start - Income_Growth_From_Start,

    Mortgage_Adjusted_HPI = Yearly_Average_HPI * Average_Mortgage_Rate
  )

glimpse(feature_data)

knitr::kable(
  head(feature_data),
  caption = "Preview of feature engineered dataset"
)
```

# Feature Descriptions

The new features are:

```{r feature-description-table}
feature_descriptions <- tibble(
  Feature = c(
    "HPI_to_Income_Ratio",
    "HPI_per_100k_Income",
    "Affordability_Index",
    "Income_Growth_From_Start",
    "HPI_Growth_From_Start",
    "Affordability_Gap",
    "Mortgage_Adjusted_HPI"
  ),
  Meaning = c(
    "Yearly average HPI divided by median household income",
    "HPI scaled per $100,000 of median household income",
    "HPI-to-income ratio compared to the first year in the dataset",
    "Median household income growth compared to the first year",
    "HPI growth compared to the first year",
    "Difference between HPI growth and income growth",
    "HPI multiplied by average mortgage rate"
  ),
  Interpretation = c(
    "Higher value suggests housing prices are higher relative to income",
    "Higher value suggests worse affordability relative to a fixed income scale",
    "Values above 1 suggest affordability worsened compared to the first year",
    "Values above 1 mean income increased compared to the first year",
    "Values above 1 mean HPI increased compared to the first year",
    "Positive values mean HPI grew faster than income",
    "Higher value suggests more buyer pressure from prices and interest rates together"
  )
)

knitr::kable(
  feature_descriptions,
  caption = "Feature engineering descriptions"
)
```

# Plot 1: HPI Relative to Income

```{r hpi-relative-to-income}
hpi_income_ratio_plot <- ggplot(feature_data, aes(x = Year, y = HPI_per_100k_Income)) +
  geom_line(linewidth = 1, color = "darkred") +
  geom_point(color = "darkred") +
  theme_minimal() +
  labs(
    title = "Housing Price Index Relative to Income",
    subtitle = "Higher values suggest worsening affordability relative to income",
    x = "Year",
    y = "HPI per $100,000 of Median Household Income"
  )

hpi_income_ratio_plot
```

# Plot 2: Relative Affordability Index

```{r affordability-index-plot}
affordability_index_plot <- ggplot(feature_data, aes(x = Year, y = Affordability_Index)) +
  geom_line(linewidth = 1, color = "purple") +
  geom_point(color = "purple") +
  geom_hline(yintercept = 1, linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "Relative Housing Affordability Index",
    subtitle = "Values above 1 mean affordability worsened compared to the first year",
    x = "Year",
    y = "Affordability Index"
  )

affordability_index_plot
```

# Plot 3: HPI Growth vs. Income Growth

```{r hpi-growth-vs-income-growth}
growth_data <- feature_data %>%
  select(Year, Income_Growth_From_Start, HPI_Growth_From_Start) %>%
  pivot_longer(
    cols = -Year,
    names_to = "Variable",
    values_to = "Growth_From_Start"
  )

growth_comparison_plot <- ggplot(growth_data, aes(x = Year, y = Growth_From_Start, color = Variable)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "HPI Growth Compared to Income Growth",
    subtitle = "Both variables are indexed to the first year in the dataset",
    x = "Year",
    y = "Growth Compared to First Year",
    color = "Variable"
  )

growth_comparison_plot
```

# Plot 4: Affordability Gap

```{r affordability-gap-plot}
affordability_gap_plot <- ggplot(feature_data, aes(x = Year, y = Affordability_Gap)) +
  geom_line(linewidth = 1, color = "orange") +
  geom_point(color = "orange") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "Affordability Gap",
    subtitle = "Positive values mean HPI grew faster than income",
    x = "Year",
    y = "HPI Growth - Income Growth"
  )

affordability_gap_plot
```

# Plot 5: Mortgage-Adjusted HPI

```{r mortgage-adjusted-hpi}
mortgage_adjusted_hpi_plot <- ggplot(feature_data, aes(x = Year, y = Mortgage_Adjusted_HPI)) +
  geom_line(linewidth = 1, color = "darkgreen") +
  geom_point(color = "darkgreen") +
  theme_minimal() +
  labs(
    title = "Mortgage-Adjusted Housing Pressure",
    subtitle = "Combines HPI and mortgage rates into one pressure indicator",
    x = "Year",
    y = "HPI x Mortgage Rate"
  )

mortgage_adjusted_hpi_plot
```

# Scatter Plot: Income vs. HPI with Regression Line

```{r income-vs-hpi-regression}
income_hpi_scatter_plot <- ggplot(
  feature_data,
  aes(x = Median_Household_Income, y = Yearly_Average_HPI)
) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  theme_minimal() +
  labs(
    title = "Median Household Income vs. Housing Price Index",
    subtitle = "Regression line shows the linear trend between income and HPI",
    x = "Median Household Income ($)",
    y = "Yearly Average HPI"
  )

income_hpi_scatter_plot
```

# Save Feature Engineered Dataset

```{r save-feature-data}
write_csv(feature_data, "feature_engineered_affordability_data.csv")

cat("Saved feature engineered dataset as: feature_engineered_affordability_data.csv\n")
```

# Written Summary

For feature engineering, we created several affordability-related variables using the cleaned housing price, income, and mortgage rate datasets. The main feature is the HPI-to-income ratio, which compares housing price index values to median household income. Since HPI is an index rather than an actual dollar home price, this ratio is used as a relative affordability indicator instead of a literal measure of how many years of income are needed to buy a house.

A larger affordability ratio or affordability index suggests that housing prices are increasing faster than income, which may indicate worsening affordability over time. We also created growth-based features to compare HPI growth and income growth relative to the first year in the dataset. Finally, we created a mortgage-adjusted HPI feature to combine the effect of housing price growth and mortgage interest rates.

```{r final-summary}
cat("Feature Engineering Summary\n\n")
cat("Created HPI-to-income ratio as a relative affordability indicator.\n")
cat("Created HPI per $100,000 of income for easier interpretation.\n")
cat("Created an affordability index relative to the first year in the dataset.\n")
cat("Created growth features comparing HPI growth and income growth.\n")
cat("Created an affordability gap feature to show whether HPI grew faster than income.\n")
cat("Created a mortgage-adjusted HPI pressure feature.\n")
cat("Skipped regional indicators because the current datasets are statewide California data.\n")
```
