---
title: "price_trends"
author: "Abigail Wong"
date: "2026-05-02"
output:
  pdf_document: default
---

```{r}
#install.packages("tidyverse")
#install.packages("ggplot2")
library(tidyverse)
library(ggplot2)
yearly_data <- read.csv("/Users/abigailwong/Desktop/ddsc_spring/CASTHPI_yearly_cleaned.csv")

glimpse(yearly_data)

ggplot(yearly_data, aes(x = Year, y = Yearly_Average_HPI)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  scale_y_continuous(limits = c(0, 920)) +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal() +
  labs(
    x = "Year",
    y = "Yearly Average HPI",
    fill = "Yearly_Average_HPI",
    title = "Yearly Average HPI between 1984 and 2023",
  )


ggplot(yearly_data, aes(x = Year, y = Lowest_Quarterly_HPI)) +
  geom_bar(stat = "identity", fill = "coral") +
  scale_y_continuous(limits = c(0, 920)) +
  theme_minimal() +
  labs(
    x = "Year", 
    y = "Lowest Quarterly HPI",
    fill = "Lowest_Quarterly_HPI",
    title = "Lowest Quarterly HPI between 1984 and 2023"
  )


ggplot(yearly_data, aes(x = Year, y = Highest_Quarterly_HPI)) +
  geom_bar(stat = "identity", fill = "darkseagreen") +
  scale_y_continuous(limits = c(0, 920)) +
  theme_minimal() +
  labs(
    x = "Year", 
    y = "Higheset Quarterly HPI",
    fill = "Highest_Quarterly_HPI",
    title = "Highest Quarterly HPI between 1984 and 2023"
  )
```

Summary: After looking at this data, we can see that the Housing Price Index 
(HPI) has more than tripled when comparing prices around 1990 to 2021 through 
2023. This tells us that housing has become more expensive over time. Between
the years 2005 - 2007, housing prices increased due to the financial crisis. 



## Visualization: Present findings through graphs

Make a correlation heatmap showing correlation coefficients between income, interest rates, and the price index to identify which variable(s) have the strongest effect on housing prices before models are run. 


income, mortgage (not interest rates), price


```{r}
income_mortgage_price <- read.delim("/Users/abigailwong/income_mortgage_price.csv",
                      header = FALSE, sep = "\t",
                      col.names = c("date", "income", "price", "mortgage", "association"))

income_mortgage_price <- income_mortgage_price |>
  filter(association == 1)

association_data <- income_mortgage_price |>
  inner_join(income_mortgage_price, by = "price", relationship= "many-to-many")

prep_data <- association_data |>
  pivot_longer(
    cols = -price,,
    names_to = "Price",
    values_to = "Strongest Effect"
  )

heatmap <- prep_data |>
  group_by(price, association) |>
  summarize(mean_strongest_effect = mean(association), .groups = "drop")

ggplot(heatmap, aes(x = association, y = price, fill = mean_strongest_effect)) +
  geom_tile() +
  scale_fill_gradient(low = "blue", high = "red") +
  labs(
    x = "Association",
    y = "Price",
    fill = "Strongest Effect"
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.text = element_text(size = 14),
    axis.text.y = element_text(size = 14)
  )
```





