---
title: " Correlation, Regression, and Noise Analysis"
author: "Harjot Singh"
date: "2026-05-03"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
  pdf_document: default
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

options(scipen = 999)
```

# Purpose

This R Markdown file adds the missing parts of Step 2 without repeating the team's existing graphs.

The team already has:

1. A normalized mortgage rate vs. HPI time-series graph.
2. An income vs. housing prices dual-axis time-series graph.

This file focuses on the remaining Step 2 work:

- Scatter plots with regression lines
- Correlation matrix
- Correlation heatmap
- Linear regression models
- R-squared comparison
- Actual vs. predicted HPI plot
- Residual/noise plot
- RMSE and MAE error metrics
- HPI yearly range as another noise/variation check

# Load Data

```{r load-data}


required_files <- c(
  "CASTHPI_yearly_cleaned.csv",
  "cleaned_income_data.csv",
  "cleaned_mortgage_rates_annual_avg.csv"
)

missing_files <- required_files[!file.exists(required_files)]

if (length(missing_files) > 0) {
  stop(
    paste(
      "Missing file(s):",
      paste(missing_files, collapse = ", "),
      "\nCurrent working directory is:",
      getwd(),
      "\nMove this Rmd file into the same folder as the CSV files or set the working directory correctly."
    )
  )
}

hpi_data <- read_csv("CASTHPI_yearly_cleaned.csv")

income_data <- read_csv("cleaned_income_data.csv") %>%
  rename(Year = Date)

mortgage_data <- read_csv(
  "cleaned_mortgage_rates_annual_avg.csv",
  skip = 7,
  col_names = c(
    "Year",
    "Average_Mortgage_Rate",
    "Lowest_Mortgage_Rate",
    "Highest_Mortgage_Rate",
    "Number_Of_Weeks"
  )
)

glimpse(hpi_data)
glimpse(income_data)
glimpse(mortgage_data)
```

# Clean and Merge Data

```{r clean-and-merge}
hpi_data <- hpi_data %>%
  mutate(
    Year = as.numeric(Year),
    Yearly_Average_HPI = as.numeric(Yearly_Average_HPI),
    Lowest_Quarterly_HPI = as.numeric(Lowest_Quarterly_HPI),
    Highest_Quarterly_HPI = as.numeric(Highest_Quarterly_HPI)
  )

income_data <- income_data %>%
  mutate(
    Year = as.numeric(Year),
    Median_Household_Income = as.numeric(Median_Household_Income)
  )

mortgage_data <- mortgage_data %>%
  mutate(
    Year = as.numeric(Year),
    Average_Mortgage_Rate = as.numeric(Average_Mortgage_Rate),
    Lowest_Mortgage_Rate = as.numeric(Lowest_Mortgage_Rate),
    Highest_Mortgage_Rate = as.numeric(Highest_Mortgage_Rate),
    Number_Of_Weeks = as.numeric(Number_Of_Weeks)
  )

combined_data <- hpi_data %>%
  left_join(income_data, by = "Year") %>%
  left_join(mortgage_data, by = "Year") %>%
  mutate(
    HPI_Range = Highest_Quarterly_HPI - Lowest_Quarterly_HPI
  ) %>%
  drop_na()

glimpse(combined_data)

knitr::kable(
  head(combined_data),
  caption = "Preview of merged housing, income, and mortgage dataset"
)
```

# Scatter Plots with Regression Lines

These plots are different from the team's existing time-series graphs. Instead of showing variables over time, these show the direct relationship between the predictor variable and HPI.

## Median Income vs. HPI

```{r income-vs-hpi-scatter}
income_vs_hpi_scatter <- ggplot(
  combined_data,
  aes(x = Median_Household_Income, y = Yearly_Average_HPI)
) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  theme_minimal() +
  labs(
    title = "Median Income vs. Housing Price Index",
    x = "Median Household Income ($)",
    y = "Yearly Average HPI"
  )

income_vs_hpi_scatter
```

## Mortgage Rate vs. HPI

```{r mortgage-vs-hpi-scatter}
mortgage_vs_hpi_scatter <- ggplot(
  combined_data,
  aes(x = Average_Mortgage_Rate, y = Yearly_Average_HPI)
) +
  geom_point(color = "darkgreen") +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  theme_minimal() +
  labs(
    title = "Mortgage Rate vs. Housing Price Index",
    x = "Average 30-Year Mortgage Rate (%)",
    y = "Yearly Average HPI"
  )

mortgage_vs_hpi_scatter
```

# Correlation Matrix

The correlation matrix shows how strongly the variables move together linearly.

Values close to `1` show a strong positive relationship.  
Values close to `-1` show a strong negative relationship.  
Values close to `0` show a weak linear relationship.

```{r correlation-matrix}
correlation_data <- combined_data %>%
  select(
    Yearly_Average_HPI,
    Median_Household_Income,
    Average_Mortgage_Rate,
    HPI_Range
  )

cor_matrix <- cor(correlation_data, use = "complete.obs")

knitr::kable(
  round(cor_matrix, 3),
  caption = "Correlation matrix for HPI, income, mortgage rate, and HPI range"
)
```

## Correlation Heatmap

```{r correlation-heatmap}
cor_df <- as.data.frame(as.table(cor_matrix))

correlation_heatmap <- ggplot(cor_df, aes(x = Var1, y = Var2, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = round(Freq, 2)), size = 4) +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  theme_minimal() +
  labs(
    title = "Correlation Matrix Heatmap",
    x = "",
    y = "",
    fill = "Correlation"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

correlation_heatmap
```

# Linear Regression and R-Squared

R-squared tells us how much of the variation in HPI is explained by each model.

This section compares three models:

1. HPI predicted by median income
2. HPI predicted by mortgage rate
3. HPI predicted by both median income and mortgage rate

```{r linear-regression}
model_income <- lm(
  Yearly_Average_HPI ~ Median_Household_Income,
  data = combined_data
)

model_mortgage <- lm(
  Yearly_Average_HPI ~ Average_Mortgage_Rate,
  data = combined_data
)

model_full <- lm(
  Yearly_Average_HPI ~ Median_Household_Income + Average_Mortgage_Rate,
  data = combined_data
)

r2_table <- tibble(
  Model = c(
    "HPI predicted by Median Income",
    "HPI predicted by Mortgage Rate",
    "HPI predicted by Median Income and Mortgage Rate"
  ),
  R_squared = c(
    summary(model_income)$r.squared,
    summary(model_mortgage)$r.squared,
    summary(model_full)$r.squared
  ),
  Adjusted_R_squared = c(
    summary(model_income)$adj.r.squared,
    summary(model_mortgage)$adj.r.squared,
    summary(model_full)$adj.r.squared
  )
)

knitr::kable(
  r2_table,
  digits = 4,
  caption = "R-squared comparison of linear regression models"
)
```

## Full Model Summary

```{r full-model-summary}
summary(model_full)
```

# Actual vs. Predicted HPI

This plot shows how close the combined regression model gets to the real HPI values.

```{r actual-vs-predicted}
combined_data <- combined_data %>%
  mutate(
    Predicted_HPI = predict(model_full, newdata = combined_data),
    Residual = Yearly_Average_HPI - Predicted_HPI
  )

actual_vs_predicted_plot <- ggplot(combined_data, aes(x = Year)) +
  geom_line(aes(y = Yearly_Average_HPI, color = "Actual HPI"), linewidth = 1) +
  geom_line(aes(y = Predicted_HPI, color = "Predicted HPI"), linewidth = 1, linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "Actual HPI vs. Predicted HPI",
    x = "Year",
    y = "House Price Index",
    color = ""
  )

actual_vs_predicted_plot
```

# Residual Plot and Noise Check

Residuals are prediction errors.

If the residuals are randomly spread around `0`, the model is doing a better job. If there is a clear pattern, then the model is missing some important factor or the relationship is not fully linear.

```{r residual-plot}
residual_plot <- ggplot(combined_data, aes(x = Year, y = Residual)) +
  geom_point(color = "purple") +
  geom_line(color = "purple") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "Residual Plot: Model Error Over Time",
    x = "Year",
    y = "Residual"
  )

residual_plot
```

# Error Metrics

RMSE and MAE measure prediction error.

Lower values mean less error.

```{r error-metrics}
rmse <- sqrt(mean(combined_data$Residual^2))
mae <- mean(abs(combined_data$Residual))

error_table <- tibble(
  Metric = c("RMSE", "MAE"),
  Value = c(rmse, mae)
)

knitr::kable(
  error_table,
  digits = 4,
  caption = "Regression model error metrics"
)
```

# HPI Range as Noise/Variation

HPI range measures the difference between the highest quarterly HPI and the lowest quarterly HPI in the same year.

A larger range means there was more movement in housing prices during that year.

```{r hpi-range-plot}
hpi_range_plot <- ggplot(combined_data, aes(x = Year, y = HPI_Range)) +
  geom_line(linewidth = 1, color = "orange") +
  geom_point(color = "orange") +
  theme_minimal() +
  labs(
    title = "Yearly HPI Range",
    subtitle = "Highest Quarterly HPI minus Lowest Quarterly HPI",
    x = "Year",
    y = "HPI Range"
  )

hpi_range_plot
```

# Final Summary

This file avoids repeating the team's existing mortgage-vs-HPI normalized time graph and income-vs-HPI dual-axis time graph. Instead, it adds the remaining parts of Step 2: correlation analysis, regression models, R-squared comparison, and noise/error analysis.

```{r final-summary}
cat("Step 2 Data Analysis Summary\n\n")
cat("This R Markdown file adds scatter plots, correlation analysis, regression R-squared comparison, and noise/error analysis.\n")
cat("The correlation matrix and heatmap help identify which variables are most linearly related to HPI.\n")
cat("The regression models compare how much income, mortgage rates, and both variables together explain HPI.\n")
cat("The residual plot, RMSE, MAE, and HPI range show how much noise or unexplained variation remains.\n")
```
